### 2023-07-11 proofs/tex/decl-typing-lemmas.tex
Context Equivalence Lemma

\newcommand{\pureSize}[1]{\ensuremath{\mathsf{pure\_size}(#1)}}
\newcommand{\eqNodes}[1]{\ensuremath{\mathsf{eq\_nodes}(#1)}}
\newcommand{\size}[1]{\ensuremath{\mathsf{size}(#1)}}

\begin{definition}
    For a tree $T$ inferring
    a declarative typing judgement 
    (i.e. the one of the shape
    $[[Γ ; Φ ⊢ v : iP]]$, 
    $[[Γ ; Φ ⊢ c : iN]]$, or 
    $[[Γ ; Φ ⊢ iN ● args ⇒> iM]]$),
    let us define a metrics $\size{T}$
    as the number of nodes in $T$ that 
    have one of these shapes
    (i.e., $[[Γ ; Φ ⊢ v : iP]]$, 
    $[[Γ ; Φ ⊢ c : iN]]$, or 
    $[[Γ ; Φ ⊢ iN ● args ⇒> iM]]$).
\end{definition}

\begin{definition}
    For a tree $T$ inferring
    a declarative typing judgement,
    let us a function $\eqNodes{T}$
    as the number of nodes in $T$ labeled with \ruleref{\ottdruleDTPEquivLabel} or 
    \ruleref{\ottdruleDTNEquivLabel}.
\end{definition}

\begin{definition}
    For a tree $T$ inferring
    a declarative typing judgement 
    let us define a metrics $\pureSize{T}$
    as the number of nodes in $T$ except 
    the ones labeled with \ruleref{\ottdruleDTPEquivLabel} or 
    \ruleref{\ottdruleDTNEquivLabel}.
    In other words, $\pureSize{T} = \size{T} - \eqNodes{T}$. 
\end{definition}

\begin{lemma}[Declarative typing is preserved under context equivalence]
    Assuming $[[Γ ⊢ Φ1]]$, $[[Γ ⊢ Φ2]]$, and $[[Γ ⊢ Φ1 ≈ Φ2]]$:
    \begin{itemize}
        \item [$+$] 
            for any tree $T_1$ inferring $[[Γ ; Φ1 ⊢ v : iP]]$, 
            there exists a tree $T_2$ inferring $[[Γ ; Φ2 ⊢ v : iP]]$ such that 
            $\pureSize{T_2} = \pureSize{T_1}$.
        \item [$-$] 
            for any tree $T_1$ inferring $[[Γ ; Φ1 ⊢ c : iN]]$, 
            there exists a tree $T_2$ inferring $[[Γ ; Φ2 ⊢ c : iN]]$ such that 
            $\pureSize{T_2} = \pureSize{T_1}$.
        \item [$\bullet$] 
            for any tree $T_1$ inferring $[[Γ ; Φ1 ⊢ iN ● args ⇒> iM]]$, 
            there exists a tree $T_2$ inferring $[[Γ ; Φ2 ⊢ iN ● args ⇒> iM]]$ such that 
            $\pureSize{T_2} = \pureSize{T_1}$.
    \end{itemize}
\end{lemma}
\begin{proof}
    Let us prove it by induction on a pair $(\pureSize{T_1}, \eqNodes{T_1})$.
    Let us consider the last rule applied in $T_1$ (i.e., its root node).
    \begin{caseof}
        \item \ruleref{\ottdruleDTVarLabel}\\
            Then we are proving 
            that $[[Γ ; Φ1 ⊢ x : iP]]$ implies $[[Γ ; Φ2 ⊢ x : iP]]$.
            By inversion, $[[x : iP ∊ Φ1]]$, and 
            since $[[Γ ⊢ Φ1 ≈ Φ2]]$, $[[x : iP' ∊ Φ2]]$ for some $[[iP']]$ 
            such that $[[Γ ⊢ iP ≈ iP']]$.
            Then we infer $[[Γ ; Φ2 ⊢ x : iP']]$ by \ruleref{\ottdruleDTVarLabel},
            and next, $[[Γ ; Φ2 ⊢ x : iP]]$ by \ruleref{\ottdruleDTPEquivLabel}.

            Notice that 
            to infer $T_1$, only \ruleref{\ottdruleDTVarLabel} was applied, 
            to infer $T_2$, only \ruleref{\ottdruleDTVarLabel} and
            \ruleref{\ottdruleDTPEquivLabel} were applied.
            This way, $\pureSize{T_1} = \pureSize{T_2} = 1$.

        \item For \ruleref{\ottdruleDTThunkLabel},
              \ruleref{\ottdruleDTPAnnotLabel}, 
              \ruleref{\ottdruleDTTLamLabel},
              \ruleref{\ottdruleDTReturnLabel},
              \ruleref{\ottdruleDTNAnnotLabel}, and
              the proof is analogous. We
              apply the induction hypothesis to the premise of the rule
              to substitute $[[Φ1]]$ for $[[Φ2]]$ in it. 
              The induction is applicable as the pure size of the 
              premises is less than the pure size of the conclusion by one.
              And after that, we apply the same rule to infer the required judgement.

              By induction hypothesis, 
              for each of the premises (i.e. the children $T_1'$ of $T_1$),
              the tree $T_2'$ inferring the corresponding judgement
              (with $[[Φ1]]$ substituted for $[[Φ2]]$) has the same pure size,
              i.e., $\pureSize{T_1'} = \pureSize{T_2'}$.
              For each of these rules, the pure size of $T_1$ 
              and the pure size $T_2$ is a sum of the pure sizes of their children
              plus one. This way, 
              $ 
                \pureSize{T_1} =
                1 + \sum_{T'_1}{\pureSize{T'_1}} =
                1 + \sum_{T'_2}{\pureSize{T'_2}} =
                \pureSize{T_2} 
              $.

        \item \ruleref{\ottdruleDTPEquivLabel} and \ruleref{\ottdruleDTNEquivLabel}
            In these cases, the induction hypothesis is also applicable to the premise:
            although the pure size of the premise is the same as the pure size of the conclusion,
            but the number of equivalence nodes is less by one.
            After this note, the proof is analogous to the previous case.

        \item \ruleref{\ottdruleDTtLamLabel}
            Then we are proving that 
            $[[Γ ; Φ1 ⊢ λx:iP.c : iP → iN]]$ implies $[[Γ ; Φ2 ⊢ λx:iP.c : iP → iN]]$.
            Analogously to the previous case, 
            we apply the induction hypothesis to the
            equivalent contexts $[[Γ ⊢ Φ1, x:iP ≈ Φ2, x:iP]]$
            and the premise $[[Γ ; Φ1, x:iP ⊢ c : iN]]$
            inferred by $T'_1$ to obtain
            $T'_2$ inferring $[[Γ ; Φ2, x:iP ⊢ c : iN]]$.
            Then we construct $T_2$ inferring $[[Γ ; Φ2 ⊢ λx:iP.c : iP → iN]]$ by \ruleref{\ottdruleDTtLamLabel}
            and notice that $\pureSize{T_1} = 1 + \pureSize{T'_1} = 1 + \pureSize{T'_2} = \pureSize{T_2}$.

@@

            The pure size of the resulting tree $T_2$ and of $T_1$ is
            one plus the sum of sizes of the premise trees.
            The premises of $T_1$ and of $T_2$ have the same pure size by 
            the induction hypothesis. This way, $\pureSize{T_1} = \pureSize{T_2}$.

@@

            It is left to show the uniqueness of $[[Γ ; Φ2 ⊢ iM ● args ⇒> ↑iQ]]$.

@@

        \item \ruleref{\ottdruleDTUnpackLabel}, \ruleref{\ottdruleDTNAnnotLabel}, and \ruleref{\ottdruleDTNEquivLabel}
            are proved in the same way.

        \item \ruleref{\ottdruleDTEmptyAppLabel}
            Then we are proving that 
            $[[Γ ; Φ1 ⊢ iN ● · ⇒> iN']]$ (inferred by \ruleref{\ottdruleDTEmptyAppLabel})
            implies $[[Γ ; Φ2 ⊢ iN ● · ⇒> iN']]$,
            and if there is only one $[[iN']]$ (up-to-equivalence in $[[Γ]]$) 
            that satisfies $[[Γ ; Φ1 ⊢ iN ● · ⇒> iN']]$,
            then there is only one $[[iN']]$ satisfying $[[Γ ; Φ2 ⊢ iN ● · ⇒> iN']]$.

@@

            To show the uniqueness $[[Γ ; Φ2 ⊢ iN ● · ⇒> iN' uniq]]$, 
            let us assume $[[Γ ; Φ2 ⊢ iN ● · ⇒> iN'']]$.
            Then the only rule applicable to infer it is \ruleref{\ottdruleDTEmptyAppLabel},
            which, by inversion, means $[[Γ ⊢ iN ≈ iN'']]$, and by transitivity, 
            $[[Γ ⊢ iN' ≈ iN'']]$.

@@

        \item \ruleref{\ottdruleDTForallAppLabel}
    \end{caseof}

>>>>

\newcommand{\pureSize}[1]{\ensuremath{\mathsf{pure\_size}(#1)}}
\newcommand{\metric}[1]{\ensuremath{\mathsf{metric}(#1)}}
\newcommand{\eqNodes}[1]{\ensuremath{\mathsf{eq\_nodes}(#1)}}
\newcommand{\size}[1]{\ensuremath{\mathsf{size}(#1)}}
\newcommand{\npq}[1]{\ensuremath{\mathsf{npq}(#1)}}

\begin{definition}[Number of prenex quantifiers]
    Let us define $\npq{[[iN]]}$ and $\npq{[[iP]]}$ as the number of prenex quantifiers in these types, i.e.
    \begin{itemize}
        \item [$+$] $\npq{[[∃nas.iP]]} = |nas|$, if $[[iP ≠ ∃nbs.iP']]$,
        \item [$-$] $\npq{[[∀pas.iN]]} = |pas|$, if $[[iN ≠ ∀pbs.iN']]$.
    \end{itemize}
\end{definition}

\begin{definition}[Size of a Judgement]
    \label{def:decl-typing-size}
    For a declarative typing judgement $J$
    let us define a metrics $\size{J}$ as a pair of numbers 
    in the following way:
    \begin{itemize}
        \item [$+$] $\size{[[Γ ; Φ ⊢ v : iP]]} = (\size{[[v]]}, 0)$;
        \item [$-$] $\size{[[Γ ; Φ ⊢ c : iN]]} = (\size{[[c]]}, 0)$;
        \item [$\bullet$] $\size{[[Γ ; Φ ⊢ iN ● args ⇒> iM]]} = 
            (\size{[[args]]}, \npq{[[iN]]})$)
    \end{itemize}
    where $\size{[[v]]}$ or $\size{[[c]]}$ is the size of the 
    syntax tree of the term $[[v]]$ or $[[c]]$
    and $\size{[[args]]}$ is the sum of sizes of the terms in $[[args]]$.
\end{definition}

\begin{definition}[Number of Equivalence Nodes]
    For a tree $T$ inferring
    a declarative typing judgement,
    let us a function $\eqNodes{T}$
    as the number of nodes in $T$ labeled with \ruleref{\ottdruleDTPEquivLabel} or 
    \ruleref{\ottdruleDTNEquivLabel}.
\end{definition}

\begin{definition}[Metric]
    For a tree $T$ inferring
    a declarative typing judgement $J$,
    let us define a metric $\metric{T}$
    as a pair $(\size{J}, \eqNodes{T})$.
\end{definition}

\begin{lemma}[Declarative typing is preserved under context equivalence]
    Assuming $[[Γ ⊢ Φ1]]$, $[[Γ ⊢ Φ2]]$, and $[[Γ ⊢ Φ1 ≈ Φ2]]$:
    \begin{itemize}
        \item [$+$] 
            for any tree $T_1$ inferring $[[Γ ; Φ1 ⊢ v : iP]]$, 
            there exists a tree $T_2$ inferring $[[Γ ; Φ2 ⊢ v : iP]]$.
        \item [$-$] 
            for any tree $T_1$ inferring $[[Γ ; Φ1 ⊢ c : iN]]$, 
            there exists a tree $T_2$ inferring $[[Γ ; Φ2 ⊢ c : iN]]$.
        \item [$\bullet$] 
            for any tree $T_1$ inferring $[[Γ ; Φ1 ⊢ iN ● args ⇒> iM]]$, 
            there exists a tree $T_2$ inferring $[[Γ ; Φ2 ⊢ iN ● args ⇒> iM]]$.
    \end{itemize}
\end{lemma}
\begin{proof}
    Let us prove it by induction on the $\metric{T_1}$.
    Let us consider the last rule applied in $T_1$ (i.e., its root node).
    \begin{caseof}
        \item \ruleref{\ottdruleDTVarLabel}\\
            Then we are proving 
            that $[[Γ ; Φ1 ⊢ x : iP]]$ implies $[[Γ ; Φ2 ⊢ x : iP]]$.
            By inversion, $[[x : iP ∊ Φ1]]$, and 
            since $[[Γ ⊢ Φ1 ≈ Φ2]]$, $[[x : iP' ∊ Φ2]]$ for some $[[iP']]$ 
            such that $[[Γ ⊢ iP ≈ iP']]$.
            Then we infer $[[Γ ; Φ2 ⊢ x : iP']]$ by \ruleref{\ottdruleDTVarLabel},
            and next, $[[Γ ; Φ2 ⊢ x : iP]]$ by \ruleref{\ottdruleDTPEquivLabel}.

        \item For \ruleref{\ottdruleDTThunkLabel},
              \ruleref{\ottdruleDTPAnnotLabel}, 
              \ruleref{\ottdruleDTTLamLabel},
              \ruleref{\ottdruleDTReturnLabel}, and
              \ruleref{\ottdruleDTNAnnotLabel}
              the proof is analogous. We
              apply the induction hypothesis to the premise of the rule
              to substitute $[[Φ1]]$ for $[[Φ2]]$ in it. 
              The induction is applicable because 
              the metric of the 
              premises is less than the metric of the conclusion:
              the term in the premise is a syntactic subterm of the
              term in the conclusion.

              And after that, we apply the same rule to infer the required judgement.

        \item \ruleref{\ottdruleDTPEquivLabel} and \ruleref{\ottdruleDTNEquivLabel}
            In these cases, the induction hypothesis is also applicable to the premise:
            although the first component of the metric 
            is the same for the premise and the conclusion:
            $\size{[[Γ ; Φ ⊢ c : iN']]} = \size{[[Γ ; Φ ⊢ c : iN]]} = \size{[[c]]}$,
            the second component of the metric is less for the premise by one,
            since the equivalence rule was applied to turn the premise tree into
            $T1$.
            Having made this note, we continue the proof in the same way as in the previous case.

        \item \ruleref{\ottdruleDTtLamLabel}
            Then we are proving that 
            $[[Γ ; Φ1 ⊢ λx:iP.c : iP → iN]]$ implies $[[Γ ; Φ2 ⊢ λx:iP.c : iP → iN]]$.
            Analogously to the previous cases, 
            we apply the induction hypothesis to the
            equivalent contexts $[[Γ ⊢ Φ1, x:iP ≈ Φ2, x:iP]]$
            and the premise $[[Γ ; Φ1, x:iP ⊢ c : iN]]$
            to obtain $[[Γ ; Φ2, x:iP ⊢ c : iN]]$.
            Notice that $[[c]]$ is a subterm of $[[λx:iP.c]]$,
            i.e., the metric of the premise tree is less than the metric of the conclusion, 
            and the induction hypothesis is applicable.
            Then we infer $[[Γ ; Φ2 ⊢ λx:iP.c : iP → iN]]$ by \ruleref{\ottdruleDTtLamLabel}.

@@

@@

            It is left to show the uniqueness of $[[Γ ; Φ2 ⊢ iM ● args ⇒> ↑iQ]]$.
            Let us assume that this judgement holds for other $[[iQ']]$, 
            i.e.  there exists a tree $T_0$ inferring 
            $[[Γ ; Φ2 ⊢ iM ● args ⇒> ↑iQ']]$.
            Then notice that the induction hypothesis is applicable to
            $T_0$: the first component of the first component of $\metric{T_0}$
            is $S = \sum_{[[v]] \in [[args]]} \size{[[v]]}$, and it is less than
            the corresponding component of $\metric{T_1}$, which is
            $\size{[[let x = v(args); c]]} = 1 + \size{[[v]]} + \size{[[c]]} + S$.
            This way, $[[Γ ; Φ1 ⊢ iM ● args ⇒> ↑iQ']]$ holds by the induction hypothesis,
            but since $[[Γ ; Φ1 ⊢ iM ● args ⇒> ↑iQ uniq]]$, we have $[[Γ ⊢ iQ' ≈ iQ]]$.

@@

        \item \ruleref{\ottdruleDTUnpackLabel}, and \ruleref{\ottdruleDTNAnnotLabel}
            are proved in the same way.

        \item \ruleref{\ottdruleDTEmptyAppLabel}
            Then we are proving that 
            $[[Γ ; Φ1 ⊢ iN ● · ⇒> iN']]$ (inferred by \ruleref{\ottdruleDTEmptyAppLabel})
            implies $[[Γ ; Φ2 ⊢ iN ● · ⇒> iN']]$.

@@

@@

        \item \ruleref{\ottdruleDTForallAppLabel}
            Then we are proving that
            $[[Γ ; Φ1 ⊢ ∀pas.iN ● args ⇒> iM]]$ (inferred by \ruleref{\ottdruleDTForallAppLabel})
            implies $[[Γ ; Φ2 ⊢ ∀pas.iN ● args ⇒> iM]]$.

            By inversion, we have $[[σ]]$ such that $[[Γ ⊢ σ : pas]]$ and
            $[[Γ ; Φ1 ⊢ [σ]iN ● args ⇒> iM]]$ is inferred.
            Let us denote the inference tree as $T_1'$.
            Notice that the induction hypothesis is applicable to $T_1'$:
            $\metric{T_1'} = ((\size{[[args]]}, 0), x)$ (for some $x$) is less than 
            $\metric{T_1} = ((\size{[[args]]}, |[[pas]]|), y)$ (for some $y$),
            since $|[[pas]]| > 0$ by inversion.

            This way, by the induction hypothesis, 
            there exists a tree $T_2'$ inferring
            $[[Γ ; Φ2 ⊢ [σ]iN ● args ⇒> iM]]$.
            Notice that the premises $[[args ≠ ·]]$, $[[Γ ⊢ σ : pas]]$,
            and $[[pas ≠ ·]]$ do not depend on the variable context,
            and hold by inversion.
            Then we infer $[[Γ ; Φ2 ⊢ ∀pas.iN ● args ⇒> iM]]$ by \ruleref{\ottdruleDTForallAppLabel}.
    \end{caseof}
